\message{ !name(template.tex)}\documentclass[pageno]{jpaper}

%replace XXX with the submission number you are given from the ISCA submission site.
\newcommand{\IWreport}{2012}

\usepackage[normalem]{ulem}

\begin{document}

\message{ !name(template.tex) !offset(-3) }


\title{
A Latex template for Independent Work Reports}

\date{}
\maketitle

\thispagestyle{empty}

\begin{abstract}
This    serve as a sample you can use for independent work reports.  We provide some guidelines on content and formatting.  They are not required, but they might be helpful.
\end{abstract}

\section{Introduction}

This document is a sample you can use for formatting your independent work
report.  An example template with all bib and supporting files can be downloaded
from the IW website at {\em http://iw.cs.princeton.edu}.

\section{Base Language}

\subsection{Untyped Lambda Calculus}

There are no minimum or maximum length limits on IW reports.  
We are including this template because we think it will be helpful
for citing things properly and for including figures into formatted
text.  If you are using \LaTeX~\cite{lamport94} 
to typeset your paper, then we strongly suggest
that you start from the template available at
http://iw.cs.princeton.edu -- this
document was prepared with that template.  
If you are using a different software package to typeset your paper, 
then you can still use this document as a reasonable sample of 
how your report might look.  Table~\ref{table:formatting} is a suggestion
of some formatting guidelines, as well as being an example of how to
include a table in a Latex document.


\message{ !name(template.tex) !offset(60) }

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
